---
title: "An R Package for Fast Sampling from von Mises Fisher Distribution"
author: "Elysée Aristide Houndetoungan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    citation_package: natbib
    number_sections: true
  bookdown::pdf_book:
    citation_package: biblatex
bibliography: ["References.bib", "Packages.bib"]
biblio-style: "apalike"
link-citations: true
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{vMFdocx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context{-#over}

I build an R package named **vMF** which samples from the von Mises-Fisher distribution ($\mathcal{M}$) as [**movMF**](https://CRAN.R-project.org/package=movMF). However, unlike the [**movMF**](https://CRAN.R-project.org/package=movMF) package [@R-movMF] which also simulates and estimates mixtures of $\mathcal{M}$, **vFM** instead focuses on fast sampling from $\mathcal{M}$. The package also computes the density and the normalization constant of the von Mises-Fisher distribution.

Note that [**movMF**](https://CRAN.R-project.org/package=movMF) is more general and can be used for many other purposes. It cannot be replaced by **vMF** which is more specific. 


The von Mises Fisher distribution is used to model coordinates on a hypersphere of dimension $p \ge 2$. It can be considered as the equivalent of the normal distribution on a hypersphere. The  von Mises Fisher distribution is characterized by two parameters. The location (or mean directional) parameter $\boldsymbol{\mu}$ around which simulations from the distribution will be concentrated and the intensity parameter $\eta$ which measures the intensity of concentration of the simulations around $\boldsymbol{\mu}$. The higher $\eta$, the more the simulations are concentrated around $\boldsymbol{\mu}$. Comparing to the normal distribution, $\boldsymbol{\mu}$ is similar to the mean parameter of the normal distribution and $\dfrac{1}{\eta}$ is similar to the standard deviation.

There are several definitions of the density function of $\mathcal{M}$. In this package, the density is normalized by the uniform distribution without loss of generality. This is also the case in @mardia2009directional and @hornik2013conjugate.

Let $\mathbf{z} \sim \mathcal{M}\left(\eta,\boldsymbol{\mu}\right)$. Then,

$$f_p(\mathbf{z}|\eta, \boldsymbol{\mu}) =C_p(\eta) e^{\eta\mathbf{z}'\boldsymbol{\mu}},$$
where $\displaystyle C_p(x) = \left(\frac{x}{2}\right)^{\frac{p}{2}-1}\frac{1}{\Gamma\left(\frac{p}{2}\right)I_{\frac{p}{2}-1}(x)}$ is the normalization constant and $I_.(.)$ the Bessel function of the first kind defined by: $$\displaystyle I_{\alpha}(x) = \sum_{m=0}^{\infty}\frac{\left(\frac{x}{2}\right)^{2m+\alpha}}{m!\Gamma(m+\alpha + 1)}.$$
\noindent The normalization with respect to the uniform distribution simplifies some results. For example, $C_p(0)=1$.

## Simulation from von Mises Fisher distribution{-#sim}
The following algorithm provides a rejection sampling scheme for drawing a sample from the $\mathcal{M}$ with mean directional parameter $\boldsymbol{\mu} = (0, ... , 0, 1)$ and concentration (intensity) parameter $\eta \ge 0$ [see Section 2.1 in @hornik2014movmf]. 

* Step 1. Calculate $b$ using * Step 1. Calculate $b$ using

$$b = \dfrac{p - 1}{2\eta + \sqrt{2\eta^2 + (p - 1)^2}}.$$ Let $x_0 = (1 - b)/(1 + b)$ and $c = \eta x_0 + (p - 1)\log\left(1 - x_0^2\right)$.

* Step 2. Generate $Z \sim Beta((p-1)/2,(p-1)/2)$ and $U \sim Unif([0, 1])$ and calculate

$$W = \dfrac{1-(1+b)Z}{1-(1-b)Z}.$$

* Step 3. If

$$\eta W+(p-1)\log(1-x_0W)-c<\log(U),$$ go to step 2.

* Step 4. Generate a uniform $(d-1)$-dimensional unit vector $V$ and return

$$X =\left(\sqrt{1- W^2}V^{\prime},W\right)^{\prime}$$

The uniform ($d-1$)-dimensional unit vector $V$ can be generated by simulating $d-1$ independent
standard normal random variables and normalizing them. To get
samples from $\mathcal{M}$ with arbitrary mean direction parameter $\boldsymbol{\mu}$, $X$ is multiplied
from the left with a matrix where the first $d-1$ columns consist of unitary basis vectors of
the subspace orthogonal to $\boldsymbol{\mu}$ and the last column is equal to $\boldsymbol{\mu}$.

## Package Installation{-#vmf}
The code source of the package is written in C++ with the package **Rcpp** [@R-Rcpp]. **vMF**  is currently available on [GitHub](https://github.com/ahoundetoungan/vMF). Its installation requires to prior install **devtools** package [@R-devtools]. Moreover, Windows users should install Rtools compatible with their R version.

**vMF** package can be installed from this GitHub repos using the `install_github` function of **devtools**. All the dependencies will also be installed automatically.

```{r install, echo = TRUE, eval = FALSE}
library(devtools) 
install_github("ahoundetoungan/vMF")
```
The option `build_vignettes = TRUE` can be added if one desires to install the vignettes.

<div id="refs"></div>